---
title: "Fireflow Test Results"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    # this is important to prevent pandoc from splitting figure captions across
    # multiple lines, which results in some weird-looking plots
    pandoc_args: ["--wrap=none"]
---

``` {r echo = FALSE}
# read SVG and strip off the SVG bit on top
embed_svg <- function(path) {
  svg_lines <- readLines(path)
  out <- if (grepl("^\\s*<\\?xml", svg_lines[1])) {
    svg_lines[-1]
  } else {
    svg_lines
  }
  cat(out, sep = "\n")
}
```

# Results Overview

```{r, results = "asis", echo = FALSE, fig.height = 16, fig.width = 12}
suppressMessages(library(tidyverse))
suppressMessages(library(ggh4x))
df_machines <- read_tsv(
  snakemake@input[["machines"]],
  col_types = cols(
    dataset = "i",
    .default = "c"
  )
)

df_plot <- df_machines %>%
  mutate(repo = map_chr(filepath, \(x) str_split_1(x, "/")[[2]])) %>%
  mutate(repo = if_else(repo == "misc_repo", "zenodo", repo))

colors <- df_plot %>%
  arrange(vendor_short) %>%
  select(vendor_short, machine_short) %>%
  unique() %>%
  group_by(vendor_short) %>%
  mutate(color = if_else((cur_group_id() %% 2) == 0, "#e8e8e8", "#c6c6c6")) %>%
  pull(color)

df_plot %>%
  ggplot(aes(y = fct_rev(software_short), fill = repo)) +
  geom_bar() +
  facet_grid2(
    vendor_short + machine_short ~ .,
    scales = "free",
    switch = "y",
    space = "free",
    strip = strip_themed(
      background_y = elem_list_rect(fill = colors)
    )
  ) +
  labs(y = NULL, x = "N FCS Files") +
  theme(
    strip.text.y.left = element_text(angle = 0),
    legend.position="bottom"
  )
```

# FCS Feature Usage Overview

The FCS standards have many features which are likely not evenly used across
vendors and software. This attempts to enumerate some common (and less common)
features.

## Keyword usage

These are optional keywords in all versions in which they are valid.

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_kw_usage"]])
```

## Scaling type

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_scaling"]])
```

## CRC usage

This checks for files that have a non-zero CRC word at the end.

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_crc"]])
```

## Datasets per file

Most files will have only one dataset in them (ie a discrete FCS data unit
defined by *\$NEXTDATA*)

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_datasets"]])
```

## *DATA* schema usage

These are different methods (schema) used to encode the bytes of data in *DATA*.

Note that "uint-var" is only possible in 3.1 and 3.2, and "mixed" is only
possible in 3.2.

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_schema"]])
```

## *\$BYTEORD* usage

Checks for either big endian, little endian, or something else (ie the PDP-11
layout which is 3,4,1,2)

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_byteord"]])
```

## *ANALYSIS* segment usage

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_analysis"]])
```

## *OTHER* segment usage

These are segments that can be used for anything. Most files have zero.

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_other_seg"]])
```

## Supplemental *TEXT* segment usage

Presumably this is only needed if *TEXT* gets so big that its ending offset
exceeds 100,000,000 and thus won't fit in *HEADER*.

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_supp_text"]])
```

## Dark segment usage

This isn't really a "feature" as much a loophole. Some files have extra data
that isn't indexed by any segment offsets. In that sense these are "dark" since
they can't be seen in a standards-compliant manner. However, they are still
there and probably mean something to someone, somewhere. These are called "dark
bytes" or "dark segments" (like "dark matter").

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_dark"]])
```

# Compliance Assessment

## Parse Errors

This is a summary of all errors that `fireflow` detected when parsing files. A
description of each error (see x axis) is given below. Note that in many cases
these are aggregate errors in the sense that they indicate "at least one
keyword" (or similar) had an error.

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_errors"]])
```

### Offset errors

These are errors pertaining to offsets, ie the three offset pairs for *TEXT*,
*DATA*, and *ANALYSIS* in the *HEADER* and the *\$BEGIN/END* offsets in *TEXT*.
This also covers errors with *\$NEXTDATA* which is also an offset.

1) nc_offsets_analysis: *ANALYSIS* offsets either could not be parsed or the
   *HEADER* and *TEXT* offsets mismatched
2) nc_offsets_data: analogous to (1) for *DATA*
3) nc_offsets_data_len: *DATA* was not evenly divisible by the event width and
   was truncated to make this true.
4) nc_offsets_nextdata: *$NEXTDATA* does not actually point to the next dataset
   if it is non-zero.
5) nc_offsets_overflow_last: Final segment offset in the file exceeded the end
   of the file
6) nc_offsets_pseudoempty: At least one offset pair was empty but was written as
   `0,-1` or similar.
7) nc_offsets_supp: The supplemental *TEXT* offsets were either unparsable or
   duplicated another segment offset pair exactly.

### TEXT layout errors

These are errors pertaining to the *TEXT* segment byte layout.

Note that supplemental *TEXT* is included when evaluating these errors if it
exists. For almost all files this is not relevant.

1) nc_text_byte_pair: A key or value had an invalid byte sequence. For keys,
   means a non-ASCII byte was encountered. For values this means a non-UTF8 byte
   was encountered.
2) nc_text_delim_number: The *TEXT* segment had an even number of delimiters.
   Note that every value *TEXT* segment (regardless of escaping) must have an
   odd delimiter number.
3) nc_text_delimiter: The delimiter was not a byte from 1 to 126.
4) nc_text_non_escape: The delimiter needed to be escaped for *TEXT* to be
   parsed validly.
5) nc_text_nonstd_nonunique: A given non-stdandard key (no prepended "$") was
   encountered at least twice.
6) nc_text_std_nonunique: A given stdandard key (prepended "$") was encountered
   at least twice.
7) nc_text_token_blank_key: A blank key was encountered.
8) nc_text_token_blank_value: A blank value was encountered. This is only
   possible for escaped *TEXT* segments.
9) nc_text_token_boundary: An escaped delimiter pair was encountered next to
   a real delimiter. This is an error because it is ambiguous.
10) nc_text_token_number: The number of tokens was even. All valid *TEXT*
    must have an even number of tokens since these need to be grouped as
    key/value pairs.
11) nc_text_token_trimmed: A value which was entirely whitespace was trimmed to
    a blank value. This is only possible for unescaped *TEXT* segments.
12) nc_text_utf8: *TEXT* had invalid UTF-8 bytes. `fireflow` in these cases
    interprets *TEXT* as Latin-1 (aka ISO-8859-1).

### Keyword value standardization errors

These errors pertain to standardization of the keyword values into a
higher-level data schema specific to each FCS version.

They can be roughly grouped into general and specialized parsing errors:

#### General

1) nc_key_val_optional: At least one optional keyword could not be parsed and
   was dropped.
2) nc_key_val_scale_forced: At least one *\$PnE* keyword encoded a non-linear
   scale for a measurement that required a linear scale (either the time
   measurement or a floating-point measurement).
3) nc_key_val_scale_log: At least one *\$PnE* keyword had an invalid log scale
   value (something like `1,0`)
4) nc_key_val_timestep_invalid: *\$TIMESTEP* could not be parsed and was required.
5) nc_key_val_timestep_missing: *\$TIMESTEP* was missing and was required.
6) nc_key_val_ws_edge: At least one value had whitespace on either the left or
   right that was trimmed off to enable valid parsing of the value (usually a
   number)
7) nc_key_val_ws_inner: At least one sub-value in a comma-separated keyword
   value had whitespace between the value and the comma that needed to be
   trimmed to enable a valid parse.
8) nc_key_val_pnn_non_unique: At least one pair of *\$PnN* values were not
   unique.

#### Specialized

1) nc_parse_begindatetime_fmt: The format for *\$BEGINDATETIME* was invalid.
2) nc_parse_enddatetime_fmt: Same as (1) for *\$ENDDATETIME*.
3) nc_parse_btim_fmt: Same as (1) for *\$BTIM*.
4) nc_parse_etim_fmt: Same as (1) for *\$ETIM*.
5) nc_parse_date_fmt: Same as (1) for *\$DATE*.
6) nc_parse_last_modified_fmt: Same as (1) for *\$LAST_MODIFIED*.
7) nc_parse_spillover_names: The names for *\$SPILLOVER* were actually numbers
   which should be properly interpreted as measurement indices.
   
### Extra keyword errors

These are errors pertaining to keywords that look like standardized keywords (ie
they have a "$" in the front) yet are not really part of the FCS standard for
the file.

Note, this has a relationship to the `nc_version` error in the next section
because it depends on which FCS version the file is assumed to follow based on
a heuristic that guesses the version given the keywords present.

1) nc_extra_hyper_gate: At least one *\$Gm* keyword has an index which exceeds
   *\$GATE*.
2) nc_extra_hyper_par: At least one *\$Pn* keyword has an index which exceeds
   *\$PAR*.
3) nc_extra_other_version: At least one keywords was found which belongs to
   another version but is invalid in this file's selected version.
4) nc_extra_pseudostd: At least one keyword was found which starts with "$" but
   does not belong any FCS standard version.
5) nc_extra_temporal_optical: At least one keyword in the time measurement was
   found which should only be present on an optical measurement (ie *\$PnL*,
   *\$PnG*, etc).
   
### Misc errors

1) nc_bitmask: At least one measurement had an integer value which exceeded the
   *$PnR* bitmask.
2) nc_byteord: *$BYTEORD* did not reflect the actual layout of data in the 
   *DATA* segment. This only applies to integer layouts for FCS 2.0/3.0.
3) nc_crc_validity: The CRC existed and matched the computed CRC of the file.
4) nc_crc_word: The 8-byte CRC word was found immediately after the last segment
   of the file. It may be all zero. No attempt is made to find CRCs after 
   "other stuff" in the FCS file that are not indexed as segments (although 
   these exist in many cases). This does not apply to FCS 2.0 since CRC was
   introduced in FCS 3.0.
5) nc_int_width: The width of *\$PnB* did not match the actual width of the data
   in *DATA* and/or the *\$BYTEORD* keywords. This only applies to integer
   datatypes for FCS 2.0/3.0.
6) nc_tot: *\$TOT* did not match the actualy number of events in *DATA* (as
   computed using the sum of *\$PnB* and the length of the *DATA* segment in
   bytes.
7) nc_version: The version in *HEADER* did not match the "true" version of the
   file. See next section for more details on this error.


## Determining "True" Version

Many files have keywords that are not compatible with the version indicated in
*HEADER*; usually this is lower than the "true" version.

Note that in many cases the "true" version is somewhat arbitrary since several
will work, especially if the comparison is between 3.1 and 3.2.

To find the "true" version, start will each version and eliminate those which
drop required keywords. As a special case, 2.0 and 3.0 are eliminated if *\$PnB*
has different widths and *\$DATATYPE* is `I` (the so-called "uint-var" schema).
From whatever is left, choose the version which drops the least number of
optional keywords. In many cases this will leave at least two versions which
drop zero optional keywords each. In this case the version with the greatest
number of optional keywords is selected, or the current version is selected if
it has not been eliminated.

``` {r, results = "asis", echo = FALSE}
embed_svg(snakemake@input[["plot_version_change"]])
```

## Machine Identification

In order to construct the plots above we need to determine the machine, vendor,
and ideally it software name and version. This information is not always
available in the file itself, so sometimes this is determined based on the
publication (if it exists) alongside the file.

### Direct mapping from *\$CYT*

In the simple case, machine and vendor can be inferred directly from *\$CYT*.
Ideally, *\$CYT* would store the exact model number but instead stores the
overall product family name. For example, all Auroras are simple called "Aurora"
even though they have different laser configurations that map onto differnt
products. When possible, these machines are further deliniated using heuristics
(see next section).

The following lookup table is used to make this determination (using exact
matches):

```{r, results = "asis", echo = FALSE}
read_tsv(snakemake@input[["cyt_mapping"]], col_types = "ccc") %>%
  knitr::kable()
```

### Special cases and software determination

The above *\$CYT* lookup does not always work, so other heuristics are needed
(described below). Furthermore, heuristics are needed to find software since
the FCS standards provide no way to report the name, version, build, etc of the
software that wrote the file. In some cases, no indication of software is given
in the file at all.

These heuristics are described in depth in
`workflow/scripts/python/fcsio/compile_metadata.py` (function
`get_machine_details`).

Unless otherwise noted, machine is found using *\$CYT* using the lookup table
above.

#### Agilent

Software name and version is stored in the *#NCCreator* nonstandard keyword.

#### BD

Sometimes *\$CYT* has a nickname but these usually have the machine model listed
after.

Older machines may not have *\$CYT* at all; these usually need to be determined
manually. Machines with a Cytek upgrade will usually say something like "Cytek
DxP10", sometimes with the base machine after (FACScan, FACSCalibur, or
FACSsort).

Software is reported often in a non-standard keyword called *CREATOR*, usually
with a version.

#### BC

Software on newer machines is stored in a the nonstandard keyword *SWVER*.
Software on the Gallios is stored in *@ACQSOFTWARE*. Software on older machines
such as the CyAn and MoFlo (of varying submodels) will report their software
prepended to the operating system on the *\$SYS* keyword (like `<software> /
<OS>`).

#### Bio-Rad

Software does not appear to be stored anywhere.

#### Cytek

Machines are accurately identified in *\$CYT*. However, Aurora's are further
identified by the number of lasers, which are consistent across three different
models (3, 4, and 5 lasers) which is done by checking the nonstandard keywords
*LASERnNAME*.

Software is listed in *CREATOR* on newer machines.

#### Stratedigm

Software (including version and build) is stored in the nonstandard keywords
*SOFTWARE*.

#### Standard Biotools

TODO (this is a mess)

#### Thermo Fisher

For the Attune NxT, software does not appear to be stored anywhere.

The original Attune stores the software string in *\$CYT* and the model is
implied.
